Polynomial Time and Dependent Types

### 2023-10-24 paper/paper.tex
[ paper ] typos and a reference

With our list iterator, it is now possible to write interesting
polytime programs. For example, the example used by \citet{Hofmann03}
to demonstrate the expressivity of LFPL is insertion sort. First we
define insertion of a natural into a sorted list:
\begin{displaymath}
  \mathrm{insert} \stackrel1: \Diamond \to \tyNat \to \mathrm{IList}\,\tyNat \to \mathrm{IList}\,\tyNat
\end{displaymath}
which requires some ingenuity to write to handle the case where we
find the place to insert the item and need access to the remainder of
the list. Note that, also, the function consumes a $\Diamond$ to
construct the new element of the output list, and also that the items
in the list are themselves iterable natural numbers. This is needed to
account for the

@@

To capture the complexity class NP, we use polynomial time programs
augmented with non-determinism, as one might expect. We will not need
to reason about equality of these non-deterministic programs, so we
can represent non-deterministic choices as \emph{additive} binary
trees. We suppose a non-iterable datatype defined like so:
\begin{displaymath}
  \begin{array}{l}
    \textbf{data}\,\mathrm{ND}\,(A : \mathsf{U}) : \mathsf{U}\,\textbf{where} \\
    \quad \mathsf{return} : A \to \mathrm{ND}\,A \\
    \quad \mathsf{choice} : (\BoolTy \to \mathrm{ND}\,A) \to \mathrm{ND}\,A
  \end{array}
\end{displaymath}
The crucial point here is that the subtrees are represented as a
function $\BoolTy \to \mathrm{ND}\,A$. By the typing rules of QTT,
this means that the two branches of this function can share resources
(see the encoding of the additive product types in
\cite{atkey18qtt}). Thus, each branch of this tree is polynomial time,
not the whole tree itself.

The type $\mathrm{ND}$ supports a monad interface via the usual free
monad construction, as well as an effect
$\mathrm{flip} \stackrel1: \mathrm{ND}\,\BoolTy$ providing access to a
bit of non-deterministic information. Thus a program of type
$A \to \mathrm{ND}\,B$ in the $\sigma = 1$ fragment will be a polytime
program with access to an oracle. In the $\sigma = 0$ fragment, we can
write a function that resolves non-determinism using a list of
booleans. This function returns $\mathsf{nothing}$ if the list of
booleans is insufficient to resolve all the $\mathsf{choice}$s:
\begin{displaymath}
  \mathrm{runWithOracle} \stackrel0: \mathrm{ND}\,A \to \ListTy(\BoolTy) \to \mathrm{Maybe}\,A
\end{displaymath}
With these definitions, we can define Non-deterministic Polynomial
time as a predicate on problems:
\begin{displaymath}
  \mathrm{NP}(A,P) =
  \begin{array}[t]{@{}l}
    (f \stackrel1: \mathbf{R}(A \to \mathrm{ND}(\BoolTy)))\, \otimes \\
    \quad (a \stackrel1: A) \to \left((\mathit{bs} \stackrel1: \ListTy(\BoolTy)) \otimes (\mathrm{runWithOracle}\,(\mathbf{R}^{-1}(f)\,a)\,\mathit{bs} = \mathsf{just}\,\cTrue)\right) \Leftrightarrow P\,a
  \end{array}
\end{displaymath}
Thus, a problem is in NP if there is a non-deterministic
boolean-valued polynomial time function that has a path to returning
$\cTrue$ exactly when the input is in predicate. Moreover, it is a
quick matter of programming to see that problems in NP are closed
under the type of polytime reductions given above.

@@

By changing the computation effects supplied to a program, we can
change the complexity class. To capture the class BPP of Bounded-error
Probabilistic Polynomial time \cite{aurora}, we use a (non-iterable)
data structure representing trees of probabilistic choices, where
$\mathbb{Q}[0,1]$ is some type of (non-iterable) rationals in the
closed interval $[0,1]$:
\begin{displaymath}
  \begin{array}{l}
    \textbf{data}\,\mathrm{Dist}\,(A : \mathsf{U}) : \mathsf{U}\,\textbf{where} \\
    \quad \mathsf{return} : A \to \mathrm{Dist}\,A \\
    \quad \mathsf{choice} : \mathbb{Q}[0,1] \to (\BoolTy\ to \mathrm{Dist}\,A) \to \mathrm{Dist}\,A
  \end{array}
\end{displaymath}
As in the non-deterministic case, a function $A \to \mathrm{Dist}\,B$
in the $\sigma = 1$ fragment is a polytime probabilistic
computation. Again, the use of a function type here ensures that each
branch of the tree is constructable in polynomial time, not the whole
tree. In the $\sigma = 0$ fragment we can write a function that
computes the probability of a $\mathrm{Dist}\,\BoolTy$ computation
being true:
\begin{displaymath}
  \mathrm{probTrue} \stackrel0: \mathrm{Dist}\,\BoolTy \to \mathbb{Q}[0,1]
\end{displaymath}
We can now define the class of probabilistic polynomial time decision
problems, where the decider is allowed to make probabilistic choices
as long as it is correct with probability at least $\frac{2}{3}$:
\begin{displaymath}
  \mathrm{BPP}(A,P) =
  \begin{array}[t]{l}
    (f \stackrel1: \mathbf{R}(A \to \mathrm{Dist}(\BoolTy))) \, \otimes
    (a \stackrel1: A) \to (\mathrm{probTrue}\,(\mathbf{R}^{-1}(f)\,a) \geq \frac{2}{3}) \Leftrightarrow P\,a
  \end{array}
\end{displaymath}
Again, problems in BPP are easily seen to be closed under polytime
reductions.

Probabilistic Polynomial time has previously been considered in the
setting of implicit computational complexity by
\citet{dallago_et_al:LIPIcs.MFCS.2021.35} and \citet{LagoT15}. In both
cases, they must build probabilistic choice into the language, and
have difficulty in directly capturing the class BPP due to its
semantic nature \cite{aurora}, where the correctness of implementation
is probabilistic. With a dependently-typed host language, adding
probabilistic choice as an effect, and capturing the semantic
constraint of BPP is straightforward.

>>>>

With our list iterator, it is now possible to write interesting
polytime programs. For example, the example used by \citet{Hofmann03}
to demonstrate the expressivity of LFPL is insertion sort. First we
define insertion of a natural into a sorted list:
\begin{displaymath}
  \mathrm{insert} \stackrel1: \Diamond \to \tyNat \to \mathrm{IList}\,\tyNat \to \mathrm{IList}\,\tyNat
\end{displaymath}
which requires some ingenuity to write to handle the case where we
find the place to insert the item and need access to the remainder of
the list. Note that, also, the function consumes a $\Diamond$ to
construct the new element of the output list, and also that the items
in the list are themselves iterable natural numbers. This is needed to
account for the comparisons between elements.

@@

To capture the complexity class NP, we use polynomial time programs
augmented with non-determinism, as one might expect. We will not need
to reason about equality of these non-deterministic programs, so we
can represent non-deterministic choices as \emph{additive} binary
trees. We suppose a non-iterable datatype defined like so:
\begin{displaymath}
  \begin{array}{l}
    \textbf{data}\,\mathrm{ND}\,(A : \mathsf{U}) : \mathsf{U}\,\textbf{where} \\
    \quad \mathsf{return} : A \to \mathrm{ND}\,A \\
    \quad \mathsf{choice} : (\BoolTy \to \mathrm{ND}\,A) \to \mathrm{ND}\,A
  \end{array}
\end{displaymath}
The crucial point here is that the subtrees are represented as a
function $\BoolTy \to \mathrm{ND}\,A$. By the typing rules of QTT,
this means that the two branches of this function can share resources
(see the encoding of the additive product types by
\citet{atkey18qtt}). Thus, each branch of this tree is polynomial
time, not the whole tree itself.

The type $\mathrm{ND}$ supports a monad interface via the usual free
monad construction, as well as an effect
$\mathrm{flip} \stackrel1: \mathrm{ND}\,\BoolTy$ providing access to a
bit of non-deterministic information. Thus a program of type
$A \to \mathrm{ND}\,B$ in the $\sigma = 1$ fragment will be a polytime
program with access to an oracle. In the $\sigma = 0$ fragment, we can
write a function that resolves non-determinism using a list of
booleans. This function returns $\mathsf{nothing}$ if the list of
booleans is insufficient to resolve all the $\mathsf{choice}$s:
\begin{displaymath}
  \mathrm{runWithOracle} \stackrel0: \mathrm{ND}\,A \to \ListTy(\BoolTy) \to \mathrm{Maybe}\,A
\end{displaymath}
With these definitions, we can define Non-deterministic Polynomial
time as a predicate on problems:
\begin{displaymath}
  \mathrm{NP}(A,P) =
  \begin{array}[t]{@{}l}
    (f \stackrel1: \mathbf{R}(A \to \mathrm{ND}(\BoolTy)))\, \otimes \\
    \quad (a \stackrel1: A) \to \left((\mathit{bs} \stackrel1: \ListTy(\BoolTy)) \otimes (\mathrm{runWithOracle}\,(\mathbf{R}^{-1}(f)\,a)\,\mathit{bs} = \mathsf{just}\,\cTrue)\right) \Leftrightarrow P\,a
  \end{array}
\end{displaymath}
Thus, a problem is in NP if there is a non-deterministic
boolean-valued polynomial time function that has a path to returning
$\cTrue$ exactly when the input satisfies the predicate. Moreover, it is a
quick matter of programming to see that problems in NP are closed
under the type of polytime reductions given above.

@@

By changing the computation effects supplied to a program, we can
change the complexity class. To capture the class BPP of Bounded-error
Probabilistic Polynomial time \cite{AroraBarak09}, we use a
(non-iterable) data structure representing trees of probabilistic
choices, where $\mathbb{Q}[0,1]$ is some type of (non-iterable)
rationals in the closed interval $[0,1]$:
\begin{displaymath}
  \begin{array}{l}
    \textbf{data}\,\mathrm{Dist}\,(A : \mathsf{U}) : \mathsf{U}\,\textbf{where} \\
    \quad \mathsf{return} : A \to \mathrm{Dist}\,A \\
    \quad \mathsf{choice} : \mathbb{Q}[0,1] \to (\BoolTy \to \mathrm{Dist}\,A) \to \mathrm{Dist}\,A
  \end{array}
\end{displaymath}
As in the non-deterministic case, a function $A \to \mathrm{Dist}\,B$
in the $\sigma = 1$ fragment is a polytime probabilistic
computation. Again, the use of a function type here ensures that each
branch of the tree is constructable in polynomial time, not the whole
tree. In the $\sigma = 0$ fragment we can write a function that
computes the probability of a $\mathrm{Dist}\,\BoolTy$ computation
being true:
\begin{displaymath}
  \mathrm{probTrue} \stackrel0: \mathrm{Dist}\,\BoolTy \to \mathbb{Q}[0,1]
\end{displaymath}
We can now define the class of probabilistic polynomial time decision
problems, where the decider is allowed to make probabilistic choices
as long as it is correct with probability at least $\frac{2}{3}$:
\begin{displaymath}
  \mathrm{BPP}(A,P) =
  \begin{array}[t]{l}
    (f \stackrel1: \mathbf{R}(A \to \mathrm{Dist}(\BoolTy))) \, \otimes
    (a \stackrel1: A) \to (\mathrm{probTrue}\,(\mathbf{R}^{-1}(f)\,a) \geq \frac{2}{3}) \Leftrightarrow P\,a
  \end{array}
\end{displaymath}
Again, problems in BPP are easily seen to be closed under polytime
reductions.

Probabilistic Polynomial time has previously been considered in the
setting of implicit computational complexity by
\citet{dallago_et_al:LIPIcs.MFCS.2021.35} and \citet{LagoT15}. In both
cases, they must build probabilistic choice into the language, and
have difficulty in directly capturing the class BPP due to its
semantic nature \cite{AroraBarak09}, where the correctness of
implementation is probabilistic. With a dependently-typed host
language, adding probabilistic choice as an effect, and capturing the
semantic constraint of BPP is straightforward.
